---
title: "Connecting to the federation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting to the federation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

> Every result on this page was produced by executing the code against the
> live three-site federation. To reproduce it, run repository steps 1–4 first,
> then rebuild this site. Connection parameters are read from the gitignored
> `sites.env`; the credentials it holds are the public demo credentials.

This page logs the DataSHIELD client into the three sites of the local
federation (`aphrc`, `dgh`, `iressef`), attaches the OMOP CDM resource on each,
and explores what every site holds — using dsOMOP's *aggregate* functions, which
return only disclosure-checked metadata. No patient-level data is transferred.

## Log in and attach the OMOP resource

The login parameters (site URLs, the resource path, and the demo credentials)
are written to `sites.env` by repository steps 2–4, so we read them rather than
hard-coding anything.

```{r login, results = "hide"}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

find_up <- function(f, d = getwd()) {
  repeat {
    p <- file.path(d, f); if (file.exists(p)) return(p)
    up <- dirname(d); if (identical(up, d)) stop(f, " not found"); d <- up
  }
}
read_env <- function(path) {
  ln <- trimws(readLines(path, warn = FALSE))
  ln <- ln[nzchar(ln) & !startsWith(ln, "#")]
  e <- list()
  for (kv in regmatches(ln, regexec("^([A-Za-z_][A-Za-z0-9_]*)=(.*)$", ln)))
    if (length(kv) == 3L) e[[kv[2]]] <- kv[3]
  e
}
cfg <- read_env(Sys.getenv("SITES_ENV", unset = find_up("sites.env")))

builder <- DSI::newDSLoginBuilder()
for (S in sub("_OPAL_URL$", "", grep("_OPAL_URL$", names(cfg), value = TRUE)))
  builder$append(server = tolower(S), url = cfg[[paste0(S, "_OPAL_URL")]],
                 user = cfg$OPAL_USER, password = cfg$OPAL_PASSWORD,
                 resource = cfg$OPAL_RESOURCE_PATH,
                 driver = "OpalDriver", profile = cfg$OPAL_PROFILE)
conns <- DSI::datashield.login(logins = builder$build())

# Attach the OMOP CDM resource on every server and initialise dsOMOP.
ds.omop.connect(resource = cfg$OPAL_RESOURCE_PATH, symbol = "omop", conns = conns)
```

The client is now connected to all three sites:

```{r connected}
names(conns)
```

## What tables does each site expose?

dsOMOP discovers tables at runtime by introspecting the CDM schema — it does not
hard-code an OMOP version. Here are the tables visible on the first site:

```{r tables}
tabs <- ds.omop.tables(symbol = "omop", conns = conns)
head(tabs[[1]], 12)
```

And the columns of the `person` table:

```{r columns}
ds.omop.columns("person", symbol = "omop", conns = conns)[[1]]
```

## Federated row and person counts

`ds.omop.table.stats()` returns disclosure-checked row counts and
*distinct-person* counts for each site. The GiBleed cohort is sharded by person
into disjoint slices, so the pooled distinct-person count is the sum across
sites — and for the `person` table it reconstructs the whole cohort (2 694),
without any site exposing more than its own shard.

```{r counts}
sites <- names(conns)
tbls  <- c("person", "condition_occurrence", "drug_exposure", "measurement")

per_site <- lapply(tbls, function(tb)
  ds.omop.table.stats(tb, symbol = "omop", conns = conns)$per_site)
names(per_site) <- tbls

tabulate_field <- function(field) {
  m <- t(vapply(per_site, function(ps)
    vapply(sites, function(s) as.numeric(ps[[s]][[field]]), numeric(1)),
    numeric(length(sites))))
  colnames(m) <- sites
  cbind(m, pooled = rowSums(m))
}
```

Distinct persons per table — the `person` row is the full cohort, and the event
tables show how many of those persons have records of each kind:

```{r persons}
tabulate_field("persons")
```

Total rows per table:

```{r rows}
tabulate_field("rows")
```

## Most prevalent conditions, pooled across sites

`ds.omop.concept.prevalence()` ranks concepts by the number of distinct persons
who have them, combining the three shards into a single pooled ranking. A
concept receives a pooled count only when every shard reports it, so we show the
conditions whose pooled count is complete across all sites.

```{r prevalence}
prev <- ds.omop.concept.prevalence(
  "condition_occurrence", metric = "persons", top_n = 10,
  scope = "pooled", symbol = "omop", conns = conns)

subset(prev$pooled, !suppressed, select = c("concept_name", "n_persons"))
```

## Disconnect

```{r logout, results = "hide"}
ds.omop.disconnect(symbol = "omop", conns = conns)
DSI::datashield.logout(conns)
```

Everything above came back as aggregates. The [next page](02_extraction.html)
extracts a person-level table — still entirely server-side — and confirms it
landed as an ordinary R data frame.
