---
title: "Extracting a tailored analysis dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting a tailored analysis dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

> Every result on this page was produced by executing the code against the
> live three-site federation (repository steps 1–4). Connection parameters are
> read from the gitignored `sites.env`.

dsOMOP's plan DSL lets you request *exactly* the columns you want — and rename
them on the way out — then materialises a single person-level data frame on each
server. The table is built in place; nothing patient-level is transferred. We
then use standard `dsBaseClient` to confirm the extract has landed as an
ordinary server-side R data frame, and explore it.

## Log in and attach the OMOP resource

```{r login, results = "hide"}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

find_up <- function(f, d = getwd()) {
  repeat {
    p <- file.path(d, f); if (file.exists(p)) return(p)
    up <- dirname(d); if (identical(up, d)) stop(f, " not found"); d <- up
  }
}
read_env <- function(path) {
  ln <- trimws(readLines(path, warn = FALSE))
  ln <- ln[nzchar(ln) & !startsWith(ln, "#")]
  e <- list()
  for (kv in regmatches(ln, regexec("^([A-Za-z_][A-Za-z0-9_]*)=(.*)$", ln)))
    if (length(kv) == 3L) e[[kv[2]]] <- kv[3]
  e
}
cfg <- read_env(Sys.getenv("SITES_ENV", unset = find_up("sites.env")))

builder <- DSI::newDSLoginBuilder()
for (S in sub("_OPAL_URL$", "", grep("_OPAL_URL$", names(cfg), value = TRUE)))
  builder$append(server = tolower(S), url = cfg[[paste0(S, "_OPAL_URL")]],
                 user = cfg$OPAL_USER, password = cfg$OPAL_PASSWORD,
                 resource = cfg$OPAL_RESOURCE_PATH,
                 driver = "OpalDriver", profile = cfg$OPAL_PROFILE)
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = cfg$OPAL_RESOURCE_PATH, symbol = "omop", conns = conns)
```

## Build and execute an extraction plan

We ask for two `person` columns and rename them as we extract: `gender_concept_id`
becomes `sex` and `year_of_birth` becomes `birth_year`. The plan is executed on
each server, producing a server-side data frame bound to the symbol `D`.

```{r extract, results = "hide"}
plan <- ds.omop.plan()
plan <- ds.omop.plan.person_level(
  plan,
  tables = list(person = c(sex = "gender_concept_id", birth_year = "year_of_birth")),
  name = "demographics")

ds.omop.plan.execute(plan, out = c(demographics = "D"), symbol = "omop", conns = conns)
```

## Confirm the extract landed as a server-side data frame

The columns are exactly the two we asked for, under their new names — and no
identifier column came along:

```{r colnames}
ds.colnames("D")
```

Its dimensions, pooled across the three sites, recover the full cohort of 2 694
persons:

```{r dim}
ds.dim("D", type = "combine")
```

`ds.summary()` reports `D` as an ordinary `data.frame` — this is the point where
the OMOP layer hands off to the standard DataSHIELD toolchain:

```{r summary}
ds.summary("D")[[1]]
```

## Renaming keeps federated factor harmonisation

We renamed `gender_concept_id` to `sex`, yet dsOMOP still recognises it as a
concept column and harmonises it into a factor whose levels are coded
*identically* on every site. Renaming a concept column never silently opts it
out of this coordination — the class is `factor` and the level set is the same
union of concept ids everywhere, which is the precondition for pooling it in one
model:

```{r harmonised}
ds.class("D$sex")
ds.levels("D$sex")
```

## Explore the extracted variable

From here we use plain `dsBaseClient`. The pooled mean and quantiles of
`birth_year` combine sufficient statistics from all three sites:

```{r mean}
ds.mean("D$birth_year", type = "combine")$Global.Mean
```

```{r quantiles}
ds.quantileMean("D$birth_year", type = "combine")
```

## Disconnect

```{r logout, results = "hide"}
ds.omop.disconnect(symbol = "omop", conns = conns)
DSI::datashield.logout(conns)
```

The extract is now an ordinary federated data frame. The
[next page](03_modeling.html) fits a regression model on it with `ds.glm`.
