---
title: "Federated regression on an extracted table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Federated regression on an extracted table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

> Every result on this page was produced by executing the code against the
> live three-site federation (repository steps 1–4). Connection parameters are
> read from the gitignored `sites.env`.

Because a dsOMOP extract is just an ordinary server-side data frame, the whole
`dsBaseClient` modelling toolchain applies to it directly. And because dsOMOP
harmonises `sex` into a federation-aligned factor — coded identically on every
site, as the [previous page](02_extraction.html) showed — it is model-ready
*as-is*: `ds.glm` builds the contrast for us, so the old manual 0/1 recoding is
no longer needed. We fit the same federated GLM two ways — pooled individual-level
and study-level meta-analysis — then a logistic model the other direction. Only
the models' sufficient statistics are pooled across the three sites; no patient
row ever leaves a server.

## Log in, attach, and re-extract the demographics table

```{r login, results = "hide"}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

find_up <- function(f, d = getwd()) {
  repeat {
    p <- file.path(d, f); if (file.exists(p)) return(p)
    up <- dirname(d); if (identical(up, d)) stop(f, " not found"); d <- up
  }
}
read_env <- function(path) {
  ln <- trimws(readLines(path, warn = FALSE))
  ln <- ln[nzchar(ln) & !startsWith(ln, "#")]
  e <- list()
  for (kv in regmatches(ln, regexec("^([A-Za-z_][A-Za-z0-9_]*)=(.*)$", ln)))
    if (length(kv) == 3L) e[[kv[2]]] <- kv[3]
  e
}
cfg <- read_env(Sys.getenv("SITES_ENV", unset = find_up("sites.env")))

builder <- DSI::newDSLoginBuilder()
for (S in sub("_OPAL_URL$", "", grep("_OPAL_URL$", names(cfg), value = TRUE)))
  builder$append(server = tolower(S), url = cfg[[paste0(S, "_OPAL_URL")]],
                 user = cfg$OPAL_USER, password = cfg$OPAL_PASSWORD,
                 resource = cfg$OPAL_RESOURCE_PATH,
                 driver = "OpalDriver", profile = cfg$OPAL_PROFILE)
conns <- DSI::datashield.login(logins = builder$build())
ds.omop.connect(resource = cfg$OPAL_RESOURCE_PATH, symbol = "omop", conns = conns)

plan <- ds.omop.plan()
plan <- ds.omop.plan.person_level(
  plan,
  tables = list(person = c(sex = "gender_concept_id", birth_year = "year_of_birth")),
  name = "demographics")
plan <- ds.omop.plan.options(plan, factor_concepts = TRUE)   # on by default
ds.omop.plan.execute(plan, out = c(demographics = "D"), symbol = "omop", conns = conns)
```

## Linear model: birth year by sex

`sex` is a factor coded identically on every site, so we regress on it directly
— `ds.glm` constructs the contrast itself. The reference level is `8507` (male,
the lowest concept id), so in `birth_year ~ sex` the intercept is the mean birth
year of men and the `sex8532` term is the women-minus-men difference:

```{r gaussian}
g <- ds.glm(formula = "D$birth_year ~ D$sex", family = "gaussian",
            datasources = conns)
g$coefficients
```

## Meta-analysis: the same model, study-level

`ds.glm` above pools individual-level sufficient statistics — one model fitted
over the combined federation. `ds.glmSLMA` instead fits the model *separately*
on each site and meta-combines the per-site estimates (study-level meta-analysis),
sharing no individual-level statistic across sites. On a harmonised factor the
two routes agree:

```{r slma}
gs <- ds.glmSLMA(formula = "D$birth_year ~ D$sex", family = "gaussian",
                 datasources = conns)
gs$SLMA.pooled.ests.matrix
```

## Logistic model: sex by birth year

The harmonised factor is equally usable as a binomial *response*. Modelling
`sex ~ birth_year` estimates the log-odds of being female (`8532`, the
non-reference level) as a function of birth year:

```{r binomial}
gb <- ds.glm(formula = "D$sex ~ D$birth_year", family = "binomial",
             datasources = conns)
gb$coefficients
```

## Disconnect

```{r logout, results = "hide"}
ds.omop.disconnect(symbol = "omop", conns = conns)
DSI::datashield.logout(conns)
```

The models above are standard `dsBaseClient` fits computed on a dsOMOP extract:
the OMOP layer produced *and harmonised* the analysis dataset — including the
factor coding that made `sex` model-ready across the federation — and the
ordinary federated toolchain took it from there.
