# Exploring the catalogue {#explore}

Before extracting anything we explore the federation with dsOMOP's **aggregate**
functions. These introspect each site's OMOP schema and return only
disclosure-checked metadata and counts — no patient-level data is transferred.
This is how an analyst decides *what* to study.

## What tables does each site expose?

dsOMOP discovers tables at runtime by reading the CDM schema; it does not
hard-code an OMOP version. Here are the tables visible on the first site:

```{r tables}
tabs <- ds.omop.tables(symbol = "omop", conns = conns)
head(tabs[[1]], 15)
```

## What variables live in a table?

`ds.omop.columns()` lists a table's fields. The `person` table carries the
demographics we will use, and the event tables carry the clinical concepts:

```{r columns}
ds.omop.columns("person", symbol = "omop", conns = conns)[[1]]
ds.omop.columns("condition_occurrence", symbol = "omop", conns = conns)[[1]]
```

## How big is each table, per site and pooled?

`ds.omop.table.stats()` returns disclosure-checked **row** counts and
**distinct-person** counts for each site. Because the cohort is sharded by
person into disjoint slices, the pooled distinct-person count is the sum across
sites — and for `person` it reconstructs the whole cohort (2&nbsp;694), without
any site exposing more than its own shard.

```{r counts}
sites <- names(conns)
tbls  <- c("person", "condition_occurrence", "drug_exposure", "measurement")

per_site <- lapply(tbls, function(tb)
  ds.omop.table.stats(tb, symbol = "omop", conns = conns)$per_site)
names(per_site) <- tbls

tabulate_field <- function(field) {
  m <- t(vapply(per_site, function(ps)
    vapply(sites, function(s) as.numeric(ps[[s]][[field]]), numeric(1)),
    numeric(length(sites))))
  colnames(m) <- sites
  cbind(m, pooled = rowSums(m))
}
```

Distinct persons per table — the `person` row is the full cohort, and the event
tables show how many of those persons have records of each kind:

```{r persons}
tabulate_field("persons")
```

Total rows per table:

```{r rows}
tabulate_field("rows")
```

## Which clinical concepts are most prevalent?

`ds.omop.concept.prevalence()` ranks concepts by the number of distinct persons
who have them, fusing the three shards into one pooled ranking. A concept is
reported only when every shard contributes (otherwise it is suppressed), so the
ranking is both pooled and disclosure-safe.

The most prevalent **conditions** point straight at a question — gastrointestinal
haemorrhage and peptic ulcer are both here:

```{r prev-cond}
prev_cond <- ds.omop.concept.prevalence(
  "condition_occurrence", metric = "persons", top_n = 12,
  scope = "pooled", symbol = "omop", conns = conns)
subset(prev_cond$pooled, !suppressed, select = c("concept_name", "n_persons"))
```

And the most prevalent **drugs** are dominated by NSAIDs and analgesics — the
exposure of interest for a GI-bleed study:

```{r prev-drug}
prev_drug <- ds.omop.concept.prevalence(
  "drug_exposure", metric = "persons", top_n = 12,
  scope = "pooled", symbol = "omop", conns = conns)
subset(prev_drug$pooled, !suppressed, select = c("concept_name", "n_persons"))
```

## The study this suggests

The catalogue hands us a clinically coherent question. Gastrointestinal
haemorrhage is well represented; peptic ulcer — its best-known antecedent — is
even more common; and the cohort is saturated with NSAIDs, the classic
pharmacological trigger. So we will build a person-level frame fusing:

| Source table | Feature(s) | Role |
|---|---|---|
| `person` | `sex`, `birth_year` (→ `age`) | adjustment |
| `condition_occurrence` | `gi_bleed`, `peptic_ulcer` | outcome, key predictor |
| `drug_exposure` | `nsaid` (any of 5), `nsaid_n` (count), `diclofenac` | exposure |

The next chapter extracts exactly this, on the server, in one shot.
