# Extracting a fused person-level table {#extract}

dsOMOP's **plan DSL** lets you request *exactly* the features you want — drawn
from several OMOP tables — and materialises a single person-level data frame on
each server. The fusion happens in place; nothing patient-level is transferred.
We then hand off to standard `dsBaseClient` to confirm the extract landed as an
ordinary server-side data frame.

## The concepts to extract

These are the OMOP concept ids the catalogue pointed us to — the outcome and its
antecedent condition, and the NSAID ingredient set:

```{r concepts}
NSAID <- c(aspirin   = 19059056, celecoxib = 1118084, ibuprofen = 19078461,
           naproxen  = 1115171,  diclofenac = 1124300)
GI_BLEED     <- 192671
PEPTIC_ULCER <- 4027663
```

## Building the plan

We declare one **person-level** output that fuses three tables. Each feature is
*named* — the name on the left of `=` becomes the output column — so nothing
leaks out as an opaque `x<concept_id>` column. Two feature constructors do the
clinical work:

- `omop.feature.boolean(set)` — a single **"any of the set"** flag: `TRUE` when
  the person has a record for *any* concept in the set. `nsaid` is `TRUE` for a
  person with *any* of the five NSAIDs; `diclofenac` is the flag for one agent.
- `omop.feature.count(set)` — the **total number of records** across the set,
  giving `nsaid_n`, a continuous measure of NSAID burden.

```{r plan}
plan <- ds.omop.plan()
plan <- ds.omop.plan.person_level(
  plan, name = "study",
  tables = list(
    person = c(sex = "gender_concept_id", birth_year = "year_of_birth"),
    condition_occurrence = list(
      features = list(
        gi_bleed     = omop.feature.boolean(GI_BLEED),
        peptic_ulcer = omop.feature.boolean(PEPTIC_ULCER)),
      concept_set = c(GI_BLEED, PEPTIC_ULCER)),
    drug_exposure = list(
      features = list(
        diclofenac = omop.feature.boolean(NSAID["diclofenac"]),
        nsaid      = omop.feature.boolean(unname(NSAID)),
        nsaid_n    = omop.feature.count(unname(NSAID))),
      concept_set = unname(NSAID))))

# Harmonise concept-coded columns (e.g. sex) into federation-aligned factors.
plan <- ds.omop.plan.options(plan, factor_concepts = TRUE)
```

`ds.omop.plan.preview()` resolves the plan against the schema without
materialising data — a cheap dry run before committing:

```{r preview, results = "hide"}
invisible(ds.omop.plan.preview(plan, symbol = "omop", conns = conns))
```

## Executing on the server

`ds.omop.plan.execute()` runs the plan on each site and binds the resulting
person-level frame to the symbol `D`:

```{r execute, results = "hide"}
ds.omop.plan.execute(plan, out = c(study = "D"), symbol = "omop", conns = conns)
```

## Confirming the fused frame

The columns are exactly the named features we asked for — across three tables —
and **no identifier column came along** (dsOMOP strips person ids before the
frame ever enters the DataSHIELD session):

```{r colnames}
ds.colnames("D")
```

Its dimensions, pooled across the three sites, recover the **full cohort of
2&nbsp;694 persons** with seven feature columns:

```{r dim}
ds.dim("D", type = "combine")
```

This full-cohort count is not an accident. When a person has *no* NSAID record,
or no peptic ulcer, the fused boolean/count features for that person are filled
with **0**, not `NA` — so persons are never silently dropped from the cohort by
the cross-table join. An absent event means "did not occur", which is exactly
what a `0` encodes.

`ds.summary()` confirms `D` is an ordinary server-side `data.frame` — this is the
hand-off point from the OMOP layer to the standard DataSHIELD toolchain:

```{r summary}
ds.summary("D")[[1]]$class
ds.summary("D")[[1]]$dimensions
```

The extract is now a federated data frame. The next chapter describes it.
