# Observing the extracted variables {#observe}

The fused frame `D` is now an ordinary federated data frame, so from here we use
plain `dsBaseClient`. Every statistic below combines sufficient statistics from
all three sites; every plot is disclosure-controlled on the server before
anything is returned.

## A derived feature: age

We turn `birth_year` into `age` as of 2020 with a single server-side expression.
`ds.make()` reports back that the variable was created and passes its validity
check on every site:

```{r age}
ds.make("(2020 - D$birth_year)", "age")
```

We then confirm the values are sensible — on the first site, `age` is numeric and
its distribution sits in a plausible human range (the pooled view follows below):

```{r age-check}
ds.summary("age")[[1]]
```

The mean age sits near 62 — older than a present-day population, but expected
here. The GiBleed cohort is a frozen synthetic snapshot (Synthea-generated,
released 2019-05-25, with records ending 2019-07-03), and no person in it is born
after 1986. Because the data are fixed at that 2019 horizon we derive `age`
against a constant reference year (2020) rather than the system date: a
`Sys.Date()` value would drift on every rebuild, whereas a fixed anchor keeps the
figures in this chapter reproducible.

## Prevalence over the full cohort

Because absent events were zero-filled (Chapter \@ref(extract)), these
prevalences are over the **whole 2&nbsp;694-person cohort**, not just the persons
who happen to appear in a given event table. The outcome, gastrointestinal
haemorrhage, affects 479 persons:

```{r prev-gib}
ds.table("D$gi_bleed")$output.list$TABLES.COMBINED_all.sources_counts
```

Its antecedent, peptic ulcer, is more common (802 persons):

```{r prev-ulcer}
ds.table("D$peptic_ulcer")$output.list$TABLES.COMBINED_all.sources_counts
```

Diclofenac exposure covers 850 persons:

```{r prev-diclo}
ds.table("D$diclofenac")$output.list$TABLES.COMBINED_all.sources_counts
```

### A finding hiding in the data

The "any NSAID" flag tells us something about how the cohort was built — **every
single person** has at least one NSAID record:

```{r nsaid-saturated}
ds.table("D$nsaid")$output.list$TABLES.COMBINED_all.sources_counts
```

The flag is universal (2&nbsp;694 / 2&nbsp;694), so it carries no information for
a model — it is collinear with the intercept. This is why we also extracted
`nsaid_n` (how *many* NSAID records) and `diclofenac` (a *specific* agent): within
a uniformly-exposed cohort, the **burden** and the **choice of drug** are what can
still discriminate. Extracting all three let the data reveal this; we now model
the informative ones.

## Pooled summaries of the continuous features

```{r means}
ds.mean("age", type = "combine")$Global.Mean
ds.mean("D$nsaid_n", type = "combine")$Global.Mean
```

The pooled quantiles of age:

```{r quantiles}
ds.quantileMean("age", type = "combine")
```

## Federated plots

A pooled histogram of age across the three sites — drawn from server-side
binned counts, never individual rows:

```{r hist-age}
ds.histogram("age", type = "combine")
```

And the distribution of NSAID burden (`nsaid_n`) — right-skewed, as drug-record
counts usually are:

```{r hist-nsaid}
ds.histogram("D$nsaid_n", type = "combine")
```

A federated heatmap shows the joint density of age and NSAID burden, with the
server applying its own anonymisation to the binned surface:

```{r heatmap}
ds.heatmapPlot(x = "age", y = "D$nsaid_n", type = "combine")
```

## The key association, previewed

Cross-tabulating the outcome against its antecedent already hints at the result
the model will quantify — GI haemorrhage is markedly more frequent among persons
with a peptic-ulcer history:

```{r crosstab}
ds.table("D$gi_bleed", "D$peptic_ulcer")$output.list$TABLES.COMBINED_all.sources_counts
```

The final chapter turns this into a federated logistic regression.
