# Modelling: a federated logistic GLM {#model}

Because a dsOMOP extract is just a server-side data frame, the whole
`dsBaseClient` modelling toolchain applies to it. We fit a **federated logistic
regression** of gastrointestinal haemorrhage on demographics, NSAID exposure and
prior peptic ulcer — pooling only the model's sufficient statistics across the
three sites. No patient row ever leaves a server.

## Assembling the modelling frame

`ds.glm` needs numeric responses, so we cast the boolean features to 0/1 and bind
them — together with the harmonised `sex` factor, the derived `age`, and the
NSAID burden — into one frame `M`:

```{r assemble, results = "hide"}
ds.asNumeric("D$gi_bleed",     "gi_bleed")
ds.asNumeric("D$diclofenac",   "diclofenac")
ds.asNumeric("D$peptic_ulcer", "peptic_ulcer")

ds.dataFrame(
  c("gi_bleed", "D$sex", "age", "diclofenac", "D$nsaid_n", "peptic_ulcer"),
  newobj = "M")
```

```{r colnames-m}
ds.colnames("M")[[1]]
```

## Fitting the model (with its live training trace)

`ds.glm` fits by **iteratively re-weighted least squares**, exchanging only
aggregated sufficient statistics each round. The federation reports its progress
every iteration — this is the genuine convergence trace from the live run:

```{r fit, message = TRUE}
fit <- ds.glm(
  formula = "M$gi_bleed ~ M$sex + M$age + M$diclofenac + M$nsaid_n + M$peptic_ulcer",
  family  = "binomial", datasources = conns)
```

The model converged in `r fit$iter` iterations over the full `r fit$Nvalid`-person
cohort. Here are the pooled coefficients, with standard errors, *p*-values and
odds ratios:

```{r coefs}
round(fit$coefficients, 4)
```

## Reading the result

```{r interpret, echo = FALSE}
or   <- function(v) round(exp(fit$coefficients[v, "Estimate"]), 2)
# Return a preformatted string so knitr's inline hook prints it verbatim
# instead of re-rendering tiny scientific notation as broken superscripts.
pval <- function(v) {
  p <- fit$coefficients[v, "p-value"]
  if (p < 1e-4) formatC(p, format = "e", digits = 1) else formatC(p, format = "g", digits = 2)
}
```

The headline is unambiguous and clinically coherent: a prior history of **peptic
ulcer multiplies the odds of gastrointestinal haemorrhage by about
`r or("M$peptic_ulcer")`** (*p* ≈ `r pval("M$peptic_ulcer")`), adjusting for
everything else. **NSAID burden** (`nsaid_n`) carries a smaller but significant
signal (*p* ≈ `r pval("M$nsaid_n")`), while **age**, **sex** and the
**diclofenac** flag are not individually associated with the outcome in this
cohort — exactly the kind of mixed, honest result a real analysis produces.

That `diclofenac` and the any-NSAID flag add little is itself the story from
Chapter \@ref(observe): this is an NSAID-saturated cohort, so it is the *amount*
of exposure and the *prior ulcer*, not NSAID use per se, that move the outcome.

## The same model, study-level

`ds.glm` above pools individual-level sufficient statistics — one model over the
combined federation. `ds.glmSLMA` instead fits the model **separately on each
site** and meta-combines the per-site estimates (study-level meta-analysis),
sharing no individual-level statistic across sites. The two routes agree closely,
which is the reassurance that the pooled fit is not driven by any single site:

```{r slma, message = FALSE}
slma <- ds.glmSLMA(
  formula = "M$gi_bleed ~ M$sex + M$age + M$diclofenac + M$nsaid_n + M$peptic_ulcer",
  family  = "binomial", datasources = conns)
round(slma$SLMA.pooled.ests.matrix, 4)
```

## Disconnect

The analysis is complete, so we release the OMOP handles and log out of the
federation:

```{r logout, results = "hide"}
ds.omop.disconnect(symbol = "omop", conns = conns)
DSI::datashield.logout(conns)
```

## In summary

Starting from nothing but three independent OMOP databases, we explored their
catalogues, **fused three tables into one person-level frame on each server**,
described it with federated summaries and plots, and fitted a federated logistic
regression — recovering a clinically sensible result over the full 2&nbsp;694-person
cohort while **no patient-level data ever left a site**. The OMOP layer produced
and harmonised the analysis dataset; the ordinary DataSHIELD toolchain took it
from there.
