---
title: "Federated Analysis using DataSHIELD"
subtitle: "A live, three-site OMOP CDM workflow with dsOMOP v2"
author: "David Sarrat González, Michael Ochola"
site: bookdown::bookdown_site
documentclass: book
github-repo: APHRC-DSE/Federated-Analysis-using-DataSHIELD-I
url: https://aphrc-dse.github.io/Federated-Analysis-using-DataSHIELD-I/
description: >
  A reproducible, end-to-end federated analysis over OMOP CDM data with
  DataSHIELD and dsOMOP v2. Every figure and table on these pages is the
  genuine result of executing the code against a live three-site federation.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>",
  fig.align = "center", fig.width = 7, fig.height = 4.2, out.width = "85%"
)
options(width = 90)
```

# Overview {-}

This book is the **reproducibility companion** to the paper *"Federated Analysis
using DataSHIELD"*. It is a single, **end-to-end federated analysis** that you can
re-run from scratch: it stands up a **local three-site DataSHIELD federation** over
**OMOP CDM** data, explores what each site holds, extracts one tailored
person-level table by **fusing three OMOP tables on each server**, describes it,
and finally fits a **federated logistic regression** — all without a single
patient-level row ever leaving a site.

Every result you see was produced by **executing the code live against the three
sites** while this book was rendered. Nothing is pre-baked: the counts, the
prevalence tables, the histograms and the GLM iteration trace are the genuine
output of the run.

## Why a *simulated* federation {-}

In a real DataSHIELD study the data stays where it lives: each research centre
hosts its own server and the patient-level records **never leave the institution**
— that is the entire point. That real data therefore cannot be handed to you to
reproduce the analysis.

So this book does the next best thing. On your own machine it **simulates three
research centres** — `aphrc`, `dgh` and `iressef` — each running an independent
Opal + Rock server backed by its own OMOP CDM database. The data is fully
**synthetic** (see [The dataset](#the-dataset-ohdsi-gibleed) below) and is
**sharded by person** across the three sites, so each "centre" holds a disjoint set
of patients, exactly as three real institutions would. The analysis then runs
across all three as if they were remote — the client only ever sees
disclosure-checked aggregates.

**Part I** of this book walks you through building that federation, one step per
chapter. **Part II** is the analysis itself.

## The federation {-}

```{r federation-diagram, echo = FALSE, fig.cap = "The simulated three-site federation. One DataSHIELD client logs into three independent Opal + Rock sites (the dsOMOP omop profile); each site is backed by its own OMOP CDM PostgreSQL holding a person-disjoint shard of the GiBleed cohort."}
DiagrammeR::grViz('
digraph federation {
  graph [rankdir = TB, fontname = "Helvetica", bgcolor = "transparent", nodesep = 0.5, ranksep = 0.75]
  node  [fontname = "Helvetica", fontsize = 10]
  edge  [fontname = "Helvetica", fontsize = 9, color = "#666666"]

  client [shape = box, style = "rounded,filled", fillcolor = "#1f6feb", fontcolor = "white", margin = "0.25,0.14",
          label = "DataSHIELD client (this R session)\\nDSI · DSOpal · dsBaseClient · dsOMOPClient"]

  subgraph cluster_aphrc {
    label = "aphrc    person_id % 3 = 0"; labeljust = "l"; fontsize = 10; style = "rounded,filled"; fillcolor = "#f6f8fa"; color = "#cfd7df";
    a_opal [shape = box, style = "rounded,filled", fillcolor = "#dbeafe", label = "Opal + Rock\\nomop profile = dsOMOP image"]
    a_pg   [shape = cylinder, style = filled, fillcolor = "#fef3c7", label = "PostgreSQL\\nOMOP CDM 5.3"]
    a_opal -> a_pg [label = "resource\\nomop_demo.gibleed"]
  }
  subgraph cluster_dgh {
    label = "dgh    person_id % 3 = 1"; labeljust = "l"; fontsize = 10; style = "rounded,filled"; fillcolor = "#f6f8fa"; color = "#cfd7df";
    d_opal [shape = box, style = "rounded,filled", fillcolor = "#dbeafe", label = "Opal + Rock\\nomop profile = dsOMOP image"]
    d_pg   [shape = cylinder, style = filled, fillcolor = "#fef3c7", label = "PostgreSQL\\nOMOP CDM 5.3"]
    d_opal -> d_pg [label = "resource\\nomop_demo.gibleed"]
  }
  subgraph cluster_iressef {
    label = "iressef    person_id % 3 = 2"; labeljust = "l"; fontsize = 10; style = "rounded,filled"; fillcolor = "#f6f8fa"; color = "#cfd7df";
    i_opal [shape = box, style = "rounded,filled", fillcolor = "#dbeafe", label = "Opal + Rock\\nomop profile = dsOMOP image"]
    i_pg   [shape = cylinder, style = filled, fillcolor = "#fef3c7", label = "PostgreSQL\\nOMOP CDM 5.3"]
    i_opal -> i_pg [label = "resource\\nomop_demo.gibleed"]
  }

  client -> a_opal
  client -> d_opal [label = "login(profile = omop)"]
  client -> i_opal
}
')
```

Each site is an independent Opal + Rock stack: Opal handles authentication and the
resource catalogue, while Rock is the R server that actually runs the federated
computations. Our `omop` Rock profile bundles **dsOMOP v2**, the layer that turns a
PostgreSQL OMOP CDM into DataSHIELD-ready tables. The one person-level table we
extract is analysed *in place* on each server with standard `dsBaseClient`.

## The dataset: OHDSI GiBleed {-}

The synthetic data is the OHDSI **GiBleed** demonstration cohort (**2 694
persons**), distributed by the OHDSI community in the
[EunomiaDatasets](https://github.com/OHDSI/EunomiaDatasets) repository under the
Apache-2.0 licence. It is a small, fully synthetic OMOP CDM **v5.3** dataset built
for teaching and testing — it contains **no real patients**, which is precisely why
we can publish and re-distribute it here for a reproducibility exercise.

Step 3 downloads a pinned copy, loads it into each site's PostgreSQL, and shards it
by `person_id % 3` so each site keeps one third of the persons (and *all* of their
records). The vocabulary tables (`concept`, `vocabulary`, …) carry no `person_id`
and are replicated to every site so each can resolve concept names locally.

## The study {-}

From the catalogue exploration in Chapter&nbsp;\@ref(explore) we design a
clinically coherent question and answer it federally:

> **Is a prior history of peptic ulcer — adjusting for age, sex and NSAID
> exposure — associated with gastrointestinal haemorrhage?**

To answer it we fuse three OMOP tables into one person-level analysis frame:
`person` (demographics), `condition_occurrence` (the GI-bleed outcome and the
peptic-ulcer predictor) and `drug_exposure` (NSAID exposure). Part II follows the
natural arc of the analysis:

1. **Explore** the catalogue (Chapter \@ref(explore)) — what tables, variables
   and concepts each site exposes.
2. **Extract** (Chapter \@ref(extract)) — one server-side person-level frame
   built by fusing three tables, with named features.
3. **Observe** (Chapter \@ref(observe)) — federated summaries and `dsBaseClient`
   plots of the extracted variables.
4. **Model** (Chapter \@ref(model)) — a federated logistic GLM, with its live
   training trace, plus a study-level meta-analysis.

## Reproducing it yourself {-}

Part I builds the federation in five steps that mirror the repository's numbered
folders, then opens the live connection the analysis chapters use. Each chapter
shows the **real commands** and links the exact script that runs them:

1. **Prerequisites & easy-opal** (Chapter \@ref(setup-prerequisites)) — Docker,
   Python and R; installing the `easy-opal` CLI.
2. **The three sites** (Chapter \@ref(setup-sites)) — declaring the `aphrc`,
   `dgh`, `iressef` Opal + Rock stacks and adding the dsOMOP `omop` profile.
3. **The databases** (Chapter \@ref(setup-databases)) — a PostgreSQL OMOP CDM per
   site, seeded and sharded from GiBleed.
4. **The resources** (Chapter \@ref(setup-resources)) — the dsOMOP resource that
   points each Rock session at its database.
5. **Connecting** (Chapter \@ref(setup-connect)) — logging the DataSHIELD client
   into all three sites; this live connection feeds every chapter in Part II.

If you just want the whole thing to run, the repository README has a five-command
quick start; the chapters that follow explain what each command does and why.
