---
title: "Federated Analysis using DataSHIELD"
subtitle: "A live, three-site OMOP CDM workflow with dsOMOP v2"
author: "David Sarrat González · Michael Ochola"
date: "Rendered live against the federation on `r format(Sys.Date(), '%d %B %Y')`"
site: bookdown::bookdown_site
documentclass: book
github-repo: APHRC-DSE/Federated-Analysis-using-DataSHIELD-I
url: https://aphrc-dse.github.io/Federated-Analysis-using-DataSHIELD-I/
description: >
  A reproducible, end-to-end federated analysis over OMOP CDM data with
  DataSHIELD and dsOMOP v2. Every figure and table on these pages is the
  genuine result of executing the code against a live three-site federation.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>",
  fig.align = "center", fig.width = 7, fig.height = 4.2, out.width = "85%"
)
options(width = 90)
```

# Overview {-}

This book is a **single, end-to-end federated analysis**. It connects to a
**local three-site DataSHIELD federation** built over **OMOP CDM** data, explores
what each site holds, extracts one tailored person-level table by **fusing three
OMOP tables on the server**, describes it, and finally fits a **federated logistic
regression** — all without a single patient-level row ever leaving a site.

Every result you see was produced by **executing the code live against the three
sites** while this book was rendered. Nothing is pre-baked: the histograms, the
prevalence tables and the GLM iteration trace are the genuine output of the run.

## The federation {-}

The synthetic OHDSI **GiBleed** cohort (2&nbsp;694 persons) is **sharded by
person** across three independent sites — `aphrc`, `dgh`, `iressef` — so every
person and all of their records live on exactly one site. The client sees only
disclosure-checked aggregates; the one person-level table we extract is analysed
*in place* on each server with standard `dsBaseClient`.

```
                      DataSHIELD client (this R session)
               DSI · DSOpal · dsBaseClient · dsOMOPClient
                                  │  login(profile = "omop")
        ┌─────────────────────────┼─────────────────────────┐
   ┌────▼────┐                ┌────▼────┐                ┌────▼────┐
   │  aphrc  │                │   dgh   │                │ iressef │   Opal + Rock
   │ person  │                │ person  │                │ person  │   "omop" profile
   │  %3==0  │                │  %3==1  │                │  %3==2  │   = dsOMOP image
   └────┬────┘                └────┬────┘                └────┬────┘
        │ resource omop_demo.gibleed (format omop.dbi.db)     │
   ┌────▼────┐                ┌────▼────┐                ┌────▼────┐
   │PostgreSQL│               │PostgreSQL│               │PostgreSQL│  OMOP CDM 5.3
   └─────────┘                └─────────┘                └─────────┘
```

## The study {-}

From the catalogue exploration in Chapter&nbsp;\@ref(explore) we design a
clinically coherent question and answer it federally:

> **Is a prior history of peptic ulcer — adjusting for age, sex and NSAID
> exposure — associated with gastrointestinal haemorrhage?**

To answer it we fuse three OMOP tables into one person-level analysis frame:
`person` (demographics), `condition_occurrence` (the GI-bleed outcome and the
peptic-ulcer predictor) and `drug_exposure` (NSAID exposure). The chapters follow
the natural arc of the analysis:

1. **Explore** the catalogue (Chapter \@ref(explore)) — what tables, variables
   and concepts each site exposes.
2. **Extract** (Chapter \@ref(extract)) — one server-side person-level frame
   built by fusing three tables, with named features.
3. **Observe** (Chapter \@ref(observe)) — federated summaries and `dsBaseClient`
   plots of the extracted variables.
4. **Model** (Chapter \@ref(model)) — a federated logistic GLM, with its live
   training trace, plus a study-level meta-analysis.

## Reproducing this book {-}

Stand up the federation with the five repository steps, then render the book:

```bash
bash 1_setup/install_easy_opal.sh && source .venv/bin/activate
bash 2_opal_stacks/setup_sites.sh
bash 3_databases/setup_databases.sh
bash 4_resources/setup_resources.sh
bash 5_client/setup_client.sh
Rscript -e 'bookdown::render_book("book/index.Rmd")'
```

Connection parameters (site URLs, the resource path and the **public demo
credentials**) are read from the gitignored `sites.env` that steps 2–4 write, so
there is nothing to configure by hand.

## Connecting {-}

We log the DataSHIELD client into the three sites and attach the OMOP CDM
resource on each. This single connection stays open for the whole book — every
later chapter runs against it.

```{r login, results = "hide"}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

# Locate sites.env by walking up from the book directory to the repo root.
find_up <- function(f, d = getwd()) {
  repeat {
    p <- file.path(d, f); if (file.exists(p)) return(p)
    up <- dirname(d); if (identical(up, d)) stop(f, " not found"); d <- up
  }
}
read_env <- function(path) {
  ln <- trimws(readLines(path, warn = FALSE))
  ln <- ln[nzchar(ln) & !startsWith(ln, "#")]
  e <- list()
  for (kv in regmatches(ln, regexec("^([A-Za-z_][A-Za-z0-9_]*)=(.*)$", ln)))
    if (length(kv) == 3L) e[[kv[2]]] <- kv[3]
  e
}
cfg <- read_env(Sys.getenv("SITES_ENV", unset = find_up("sites.env")))

builder <- DSI::newDSLoginBuilder()
for (S in sub("_OPAL_URL$", "", grep("_OPAL_URL$", names(cfg), value = TRUE)))
  builder$append(server = tolower(S), url = cfg[[paste0(S, "_OPAL_URL")]],
                 user = cfg$OPAL_USER, password = cfg$OPAL_PASSWORD,
                 resource = cfg$OPAL_RESOURCE_PATH,
                 driver = "OpalDriver", profile = cfg$OPAL_PROFILE)
conns <- DSI::datashield.login(logins = builder$build())

# Attach the OMOP CDM resource on every server and initialise dsOMOP.
ds.omop.connect(resource = cfg$OPAL_RESOURCE_PATH, symbol = "omop", conns = conns)
```

The client is now connected to all three sites, with a live dsOMOP handle
(`omop`) on each:

```{r connected}
names(conns)
```
