# Connecting the DataSHIELD client {#setup-connect}

The federation is built. Now we switch hats from *server administrator* to
*analyst*: we log the DataSHIELD client into all three sites and attach the OMOP
resource on each. **This single connection stays open for the rest of the book** —
every chapter in Part II runs against the `conns` and `omop` handles created here.

## Installing the client packages

The client needs four R packages. Step 5's installer pins them (in particular
`dsOMOPClient` to its `2.0.0` tag):

```bash
Rscript 5_client/install_client.R     # or: bash 5_client/setup_client.sh
```

[`5_client/install_client.R`](https://github.com/APHRC-DSE/Federated-Analysis-using-DataSHIELD-I/blob/main/5_client/install_client.R)
pulls `DSI` and `DSOpal` from CRAN, `dsBaseClient` from the DataSHIELD repository,
and `dsOMOPClient` 2.0.0 from GitHub:

| Package | Role |
|---------|------|
| `DSI` | The DataSHIELD interface — login, assign, aggregate. |
| `DSOpal` | The Opal driver for `DSI`. |
| `dsBaseClient` | The base DataSHIELD analysis functions (`ds.*`). |
| `dsOMOPClient` | The client side of dsOMOP — `ds.omop.*`. |

## Logging in

We describe the three sites with a login builder — one `append()` per site — and
log in. The connection details are the fixed values from the previous chapters:
URLs `4808{0,1,2}`, `administrator` / `password`, the resource `omop_demo.gibleed`,
and the Rock profile **`omop`** (the one carrying dsOMOP):

```{r login, results = "hide"}
library(DSI); library(DSOpal); library(dsBaseClient); library(dsOMOPClient)

sites <- list(
  aphrc   = "http://localhost:48080",
  dgh     = "http://localhost:48081",
  iressef = "http://localhost:48082"
)

builder <- DSI::newDSLoginBuilder()
for (site in names(sites))
  builder$append(server = site, url = sites[[site]],
                 user = "administrator", password = "password",
                 resource = "omop_demo.gibleed",
                 driver = "OpalDriver", profile = "omop")

conns <- DSI::datashield.login(logins = builder$build())
```

> If you changed a port or the password back in Chapter&nbsp;\@ref(setup-sites),
> change the matching value here too.

## Attaching the OMOP resource

`datashield.login` opened the sessions; now `ds.omop.connect` attaches the OMOP CDM
resource on **every** server and initialises dsOMOP, binding a live handle to the
symbol `omop`:

```{r connect, results = "hide"}
ds.omop.connect(resource = "omop_demo.gibleed", symbol = "omop", conns = conns)
```

The client is now connected to all three sites, each with a live dsOMOP handle:

```{r connected}
names(conns)
```

That is the whole federation, live. Part II uses it to explore the catalogue,
extract a fused person-level table, describe it, and fit a federated model.
